---
title: "Visualisations of UN Sustainable Development Indicators"
author: "Michael Gusev"
date: "January 14, 2018"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
library(forcats)
library(ggthemes)
library(stringr)
options(warn=-5)
country <- read.csv("C:\\Users\\mike\\Documents\\R\\worldbank-sustainable-development-goals\\worldbank-sustainable-development-goals\\data\\sdg_country.csv")
country_series <- read.csv("C:\\Users\\mike\\Documents\\R\\worldbank-sustainable-development-goals\\worldbank-sustainable-development-goals\\data\\sdg_country_series.csv")
data <- read.csv("C:\\Users\\mike\\Documents\\R\\worldbank-sustainable-development-goals\\worldbank-sustainable-development-goals\\data\\sdg_data.csv")
footnote <- read.csv("C:\\Users\\mike\\Documents\\R\\worldbank-sustainable-development-goals\\worldbank-sustainable-development-goals\\data\\sdg_footnote.csv")
series <- read.csv("C:\\Users\\mike\\Documents\\R\\worldbank-sustainable-development-goals\\worldbank-sustainable-development-goals\\data\\sdg_series.csv")
series_time <- read.csv("C:\\Users\\mike\\Documents\\R\\worldbank-sustainable-development-goals\\worldbank-sustainable-development-goals\\data\\sdg_series_time.csv")
```

This is an exploration of World Bank data on indicators that are pertinent to the Sustainable Development Goals set out by the United Nations. In 2000, the UN released a list of targets for global human development they hoped to achieve worldwide by 2030. The topics are broad, ranging from education to poverty and environmental sustainability. More info about the SDGs is available at https://sustainabledevelopment.un.org/content/documents/21252030%20Agenda%20for%20Sustainable%20Development%20web.pdf
The data I used is available at https://data.world/worldbank/sustainable-development-goals

### Data Import and Cleanup
First task will be to understand and clean up the individual datasets.
Renaming the datasets within R for simplicity:
```{r, eval=FALSE}
View(country)
View(country_series)
View(footnote)
View(series)
View(series_time)
```
Looking through these sets, it looks like `country` and `series` are definition tables that give more information about the individual country and series keys, while `country_series`, `series_time`, and `footnote` are tables of errata that correspond to specific country & series pairs, series & year pairs, or country/series/year combinations, respectively.
```{r, eval=FALSE}
View(data)
```
The `data` set is the actual values for all the indicators in each country or region over time. We'll be doing analyses based on income group and region, so we'll want to steal those from `country` and include them in `data`:
```{r}
data <- country %>%
  select(country_code, income_group, region) %>%
  right_join(data, by="country_code")
```
We will also be comparing income group or regional values to all-world values of certain indicators, so let's adjust for that:
```{r}
data$region <- fct_expand(data$region, "World")
data$income_group <- fct_expand(data$income_group, "World")
for (i in 1:nrow(data)) {
  if (data$country_code[i] == "WLD") {
    data$region[i] <- "World"
    data$income_group[i] <- "World"
  }
}
```
### Visualising Poverty
Now we're ready to make some graphs. First off, let's work with **Goal 1: End Poverty in all its forms everywhere.** We want to get an idea of how global poverty rates have progressed over time. One great indicator is `SI.POV.DDAY`, the percentage of the population of a given country that is below the international poverty line in the given year.
Let's plot this metric as it changes over time. A graph with trend lines for every country might be too cluttered, so let's group by income level:
```{r, echo=TRUE}
world_poverty <- data %>%
  filter(indicator_code == "SI.POV.DDAY", region != "") %>%
  gather(key = "year", value = "pov_rate", X1990:X2016) %>%
  mutate(year = as.integer(str_sub(year, start=2))) %>%
  group_by(income_group, year) %>%
  summarise(median_rate = median(pov_rate, na.rm = TRUE))
world_poverty %>%
  ggplot(aes(year, median_rate, color=income_group))+
  geom_point(shape = 21, fill = "white", size = 1, stroke = 1.5)+
  geom_smooth(se=FALSE, size=2, linetype="41", method="loess")+
  scale_x_continuous(breaks=seq(1990, 2015, by=5))+
  scale_y_continuous(breaks=seq(0,85,by=10))+
  labs(y="Median Poverty Rate", x=NULL, color="Income Group",
       title="Global Poverty Trends By Income Group, 1990-2016",
       subtitle="Proportion of country population living below int'l poverty line of $1.90/day, by country income group",
       caption="Each point represents the median for the corresponding income group for each year.")+
  theme_bw()+
  theme(plot.caption = element_text(face="italic"))
```

This graph shows us a few things: that poverty is distributed among income groups about how we'd expect (with lower income countries fraught with higher poverty rates); that median poverty seems to vary drastically from year to year in high-poverty groups; that world poverty is comparable with that of lower middle income countries; and finally that global poverty rates are overall decreasing over the past three decades. One possible area to explore is the variance characteristic: what causes such variation in the low income and lower middle income groups? Is it an underlying economic instability, an effect of aggregating across many countries, or just a data acquisition problem?

### Visualising Hunger
Let's now work with another SDG, **Goal 2: End hunger, achieve food security and improved nutrition and promote sustainable agriculture.** Relevant indicators include `SN.ITK.DEFC.ZS: Prevalence of undernourishment (percent of population)` as well as prevalence of growth defects like stunting or wasting among children under 5 years of age. Here we work with undernourishment statistics.
Say we want to see global stats on undernourishment from a specific year. We want to pick a recent year in which there are as few missing values as possible, for a complete picture. Let's find it using some `purrr` magic:

```{r}
data %>%
  filter(indicator_code == "SN.ITK.DEFC.ZS", region !="") %>%
  select(country_name, X2000:X2016) %>%
  map_dbl(~sum(!is.na(.)))
```
Luckily here the data seems to be available for every year (except 2016; too recent) in countries where it is available at all. So let's choose 2014 just because:
```{r}
hunger2014 <- data %>%
  filter(indicator_code == "SN.ITK.DEFC.ZS", region != "", X2014 != "NA") %>%
  select(country_name, country_code, X2014, income_group, region) %>%
  unite("CountryRegion", region, country_code, sep = " -- ", remove=FALSE)
```
We have to create the dummy column `CountryRegion` so that the data will sort nicely into regions in the graph we create; there might be a more elegant way to do this.  
```{r}
hunger2014 %>%
  ggplot()+
  geom_col(aes(CountryRegion, X2014, fill=region))+
  scale_fill_gdocs()+
  scale_y_continuous(breaks=seq(0,50,by=10))+
  labs(x = "Countries, by Region", y = "Percentage of population deemed undernourished",
       fill = "Region", title = "World Hunger Prevalence by Region, 2014")+
  theme_bw()+
  theme(axis.text.x= element_blank(), axis.ticks = element_blank())
```

It seems grouping by region does not help us much here, as there are steep outliers in all regions. It might be useful to grab those outliers and check them against the income group, to see if there are any anomalies there. Let's include any country where at least 1 in 4 people are undernourished:
```{r}
hunger2014 %>%
  filter(X2014 >= 25) %>%
  arrange(desc(income_group)) %>%
  select(country_name, income_group)
```
There are quite a few 'lower middle income' economies and even one 'upper middle income' economy, which is somewhat surprising in mapping a metric usually connoted with underdevelopment. Further lines of inquiry might include explorations of wealth inequality, other metrics of economic and human development, and historical context to explain the problem of hunger in these countries.  

### Visualising Gender Equality in Education
One critical focus of the UN in drafting the SDGs was gender equality. As an aspiring socially just civilization, we want to ensure that people of all genders have equal access to resources worldwide. Let's look at a subtopic of **Sustainable Development Goal 4: Ensure inclusive and equitable quality education and promote lifelong learning opportunities for all.**  

The GPI, or gender parity indicator, is a simple metric that calculates the ratio of females to males in a certain group. In this case, the SDG indicator `SE.ADT.1524.LT.FM.ZS` is the GPI for youth literacy, i.e. *"the ratio of females to males ages 15-24 who can both read and write with understanding a short simple statement about their everyday life."*   

In this graph, we see how the income-group average of this indicator has progressed over time, especially highlighting the low income group. We see that while youth literacy has been quite close to equitable since 2000, females are still at a deficit in all but the highest income countries.   
```{r}
educGPI <- data %>%
  filter(indicator_code == "SE.ADT.1524.LT.FM.ZS", region != "", region != "World") %>%
  gather("year", "gpi", X2000:X2014) %>%
  mutate(year = as.integer(str_sub(year, 2))) %>%
  group_by(income_group, year) %>%
  summarise(gpi = mean(gpi, na.rm=TRUE))
ggplot(educGPI)+
  geom_col(aes(year, gpi, fill=income_group), position="dodge")+
  geom_line(data=filter(educGPI, income_group == "Low income"),aes(year, gpi))+
  geom_point(data=filter(educGPI, income_group == "Low income"),
             aes(year, gpi), shape = 21, fill = "white", size = 1, stroke = 1.5)+
  theme_bw()+
  labs(y="GPI", title = "Gender Parity in Education, 2000-2014",
       subtitle = "Gender parity index for literacy rate in youth ages 15-24, by country income group",
       fill = "Income Group", color=NULL, caption="Gender parity index for youth literacy rate is the ratio of females
       to males ages 15-24 who can both read and write with understanding
       a short simple statement about their everyday life.")+
  theme(axis.title.x=element_blank(), legend.position = "right",
        plot.caption=element_text(face="italic", hjust=0.5))
```

### Understanding Energy
In our final visualisation, we explore sustainable energy access and use throughout the world. **SDG 7** is to **ensure access to affordable, reliable, sustainable and modern energy for all.** Here we aim to understand the relationship between the 'accessible' and the 'sustainable' components of this goal, i.e. the share of a country's population that has access to electricity and the share of a country's energy consumption that comes from renewable sources. (These are indicators `EG.ELC.ACCS.ZS` and `EG.FEC.RNEW.ZS`, respectively, in the data.) We'll also throw in `NY.GDP.PCAP.CD` -- a GDP per capita indicator -- into this dataset, so we can explore the role performed by income.  
```{r}
energy2012 <- data %>%
  filter(indicator_code == "EG.ELC.ACCS.ZS" | indicator_code == "EG.FEC.RNEW.ZS" | indicator_code == "NY.GDP.PCAP.CD", region !="", region != "World") %>%
  select(X2012, indicator_code, country_name, income_group) %>%
  spread(indicator_code, X2012)
ggplot(energy2012)+
  geom_point(aes(EG.ELC.ACCS.ZS, EG.FEC.RNEW.ZS, fill=income_group), shape = 21, size = 2, stroke = 0.5)+
  labs(x="Share of population with access to electricity (%)",
       y="Share of renewable energy in total energy consumption (%)", fill="Income Group",
       title="Energy Access and Renewable Consumption as Development Markers, 2012",
       subtitle="What is the interplay between a country's renewable energy consumption and its energy accessibility?")+
  theme_bw()+
  theme(plot.subtitle = element_text(face="italic"), plot.caption=element_text(face="italic"))
```

This graph does not display any clear trends, although there seem to be two general behaviors: a cluster of high-income countries with universal electricity access but varying renewable consumption, and a rough but generally negative correlation between the two energy indicators among less-developed countries. What insight can we pull out of this?  

This might be representative of the two very different ways that renewable energy is used in pre-industrialized and post-industrialized economies. Economies that have undergone rapid development following an industrial boom now have the economic resources to expend on centralized renewable infrastructure. Meanwhile, economies that have not experienced this boom often rely on decentralized renewables such as small-scale hydro and wind to power rural homesteads. This would be easier to elucidate if this dataset had a higher granularity for different types of renewables, but that is not the case.   

Let's see whether our background knowledge regarding pre-industrialized economies holds true for this subset of the data. Is access a significant predictor of renewable consumption? Are there other, clearer economic root causes? Let's filter out high and upper middle income countries and build a simple linear model:  
```{r}
lm(EG.FEC.RNEW.ZS~EG.ELC.ACCS.ZS, energy2012 %>%
       filter(income_group == "Low income" || income_group == "Lower middle income")) %>%
  summary()
```

This gives us a highly significant estimate for the coefficient of `-0.78` correlation between the two indicators. However, the R-squared for this model is only around `0.51`, so we probably couldn't use this model to predict the relationship in other years. In a future exploration of this data, it might to do to add data from more time periods or incorporate pure economic factors to clarify renewable energy trends further.

### Conclusion
This was a simple attempt to visualise some of the factors affecting the extent to which our society is sustainable; the extent to which the human project is in tune with its environment. For the most part, what we discovered was that a global problem like world hunger or poverty cannot be fully elucidated with a simple bar graph. However, we can continually clarify the picture by sifting through data, creating visualisations, and building models.
